What to Expect with Life Expectancy?

### 2024-11-27 paper/paper.qmd
Add files via upload

---
title: "What to Expect with Life Expectancy?"
subtitle: "Examining How Status, Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
author: 
  - Manjun Zhu
thanks: "Code and data are available at: [https://github.com/Karrrrmen/Life_Expectancy](https://github.com/Karrrrmen/Life_Expectancy)."
date: today
date-format: long
abstract: "Life expectancy has seen unprecedented growth in recent decades, driven by advancements in healthcare, improved access to medical services, and reduced child mortality. However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately. This research examines the impact of a nation’s economic standing and health conditions on life expectancy, exploring nuanced variables beyond those traditionally studied. By developing a linear model, the analysis identifies key relationships between economic and healthcare indicators and their influence on population longevity. The findings aim to inform health policy and economic planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(dplyr)
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(kableExtra)
library(modelsummary)
library(ggbeeswarm)
library(ggridges)

data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

@@

This study aims to address these gaps by developing a linear model to analyze the impact of a nation’s status, healthcare factors, and economic indicators on life expectancy. The response variable is Life Expectancy, while the predictors include Status (Developed vs. Developing), Diphtheria Immunization Coverage, Government Total Expenditure, Numbers of Doctors and Nurses per 10000, and Schooling. The analysis seeks to quantify the influence of these variables, providing insights into the disparities in life expectancy and identifying factors that contribute to these differences.

@@

\begin{align*} 
\text{LifeExpectancy}_{\text{developed}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
&\ \beta_5 \text{Number of Nurses per 10000} + \\
&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
\end{align*}

\begin{align*} 
\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization} + \beta_2 \text{Schooling} + \\
&\ \beta_3 \text{Health Expenditure} + \beta_4 \text{Number of Doctors per 10000} + \\
&\ \beta_5 \text{Number of Nurses per 10000} + \\
&\ \beta_6 \text{Health Expenditure} \times \text{Number of Doctors per 10000} \times \text{Number of Nurses per 10000} + \epsilon_i
\end{align*}

@@

@@

The models were built using linear regression (LM), which is appropriate for predicting life expectancy as a continuous variable. This method captures the relationship between predictors and the response variable with interpretable coefficients. For the Developed model, predictors such as Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were selected as they are widely recognized to influence life expectancy in high-income countries. In the Developing model, we added Schooling, and both Doctors and Nurses per 10,000, reflecting healthcare access and lifestyle factors relevant to developing nations. Linear regression was chosen because it is effective for normally distributed continuous outcomes, which aligns with our dataset.

>>>>

---
title: "What to Expect with Life Expectancy?"
subtitle: "Examining How Status, Healthcare Expenditure, Education, and Medical Infrastructure Influence Life Expectancy Across Developed and Developing Regions"
author: 
  - Manjun Zhu
thanks: "Code and data are available at: [https://github.com/Karrrrmen/Life_Expectancy](https://github.com/Karrrrmen/Life_Expectancy)."
date: today
date-format: long
abstract: "Life expectancy has seen unprecedented growth in recent decades, driven by advancements in healthcare, improved access to medical services, and reduced child mortality. However, this growth has not been evenly distributed across the globe, with wealthier nations benefiting disproportionately. This research examines the impact of a nation’s economic standing and health conditions on life expectancy, exploring nuanced variables beyond those traditionally studied. By developing a linear model, the analysis identifies key relationships between economic and healthcare indicators and their influence on population longevity. The findings aim to inform health policy and economic planning, providing actionable insights to address disparities and promote equitable growth in life expectancy worldwide."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load libraries
library(dplyr)
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(kableExtra)
library(modelsummary)
library(ggbeeswarm)
library(ggridges)

# load data
data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))

@@

This study aims to examine the factors influencing life expectancy across different nations by developing a linear model that incorporates both healthcare and economic variables. The response variable in the analysis is life expectancy, while the predictors include national status (Developed vs. Developing), diphtheria immunization coverage, government healthcare expenditure as a percentage of total expenditure, and the number of doctors and nurses per 10,000 people. Additionally, schooling is considered as an important educational factor. By quantifying the effects of these predictors, the study seeks to shed light on disparities in life expectancy between developed and developing nations, and to identify key factors contributing to these differences. Through this analysis, we aim to contribute valuable insights for policy development focused on improving healthcare and economic outcomes.

@@

\begin{align*} 
\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization}_i + \beta_2 \text{Schooling}_i + \\
&\ \beta_3 \text{Health Expenditure}_i + \beta_4 \text{Number of Doctors per 10000}_i + \\
&\ \beta_5 \text{Number of Nurses per 10000}_i + \\
&\ \beta_6 \text{Health Expenditure}_i \times \text{Doctors per 10000}_i \times \text{Nurses per 10000}_i + \epsilon_i
\end{align*}
\begin{align*} 
\text{LifeExpectancy}_{\text{developing}} = &\ \beta_0 + \beta_1 \text{Diphtheria Immunization}_i + \beta_2 \text{Schooling}_i + \\
&\ \beta_3 \text{Health Expenditure}_i + \beta_4 \text{Number of Doctors per 10000}_i + \\
&\ \beta_5 \text{Number of Nurses per 10000}_i + \\
&\ \beta_6 \text{Health Expenditure}_i \times \text{Doctors per 10000}_i \times \text{Nurses per 10000}_i + \epsilon_i
\end{align*}

@@

The model assumes that the predictors have a linear relationship with the response variable, i.e., the life expectancy is linearly related to the selected features. We are interested in estimating the coefficients $\beta_1$, $\beta_2$,...,$\beta_4$ which represent the effect of each predictor and interaction term in the outcome.

@@

The linear regression model was chosen for this analysis due to its simplicity and interpretability. The coefficients from this model directly show how each predictor affects life expectancy, assuming other variables remain constant. This makes it ideal for understanding the impact of healthcare and economic factors on life expectancy in both developed and developing countries.

For the developed countries model, predictors like Diphtheria Immunization, Schooling, Total Expenditure, and the number of Doctors per 10,000 were included. These variables are well-established factors influencing life expectancy in high-income nations [@life_expectancy_who; @world_health_statistics]. In the developing countries model, additional predictors such as Schooling, and healthcare access variables like Doctors and Nurses per 10,000 were added to account for differences in healthcare infrastructure.

Linear regression is suitable for our dataset, as it assumes a normally distributed continuous response variable. While more complex models were considered, linear regression was preferred for its transparency and ease of interpretation, aligning with the goals of this analysis. Additionally, previous studies have shown that healthcare access, education, and economic factors play a significant role in determining life expectancy, particularly in developing countries [@Hanushek2016; @He2020]. Including these predictors ensures the model is built on sound reasoning and established domain knowledge.
